Remote sensing riverine gas exchange and carbon efflux in ungauged rivers from SWOT observations

### 2021-05-26 manuscript_v3.Rmd
more ms updates with supp

---
title: "Potential for remotely sensing riverine gas exchange and carbon efflux from SWOT observations"
author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##1 Introduction
The Earth's carbon cycle transports and manipulates carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes. The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports much carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a]. However en route to the oceans, this river network additionally transports and transforms sediments, nutrients, organic matter, dissolved gases, and other constituents. Organic matter and inorganic nutrients are also produced and/or consumed within rivers via ecosystem metabolism [@odumPrimaryProductionFlowing1956]. The net effect of all of this is the global river network is usually supersaturated with dissolved gases and evades substantial amounts of carbon from its water surface to the atmosphere [@colePlumbingGlobalCarbon2007a].

Current estimates of total carbon dioxide ($CO_2$) evasion to the atmosphere vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Further, in-river metabolism is increasingly understood to contribute to this flux [@hotchkissSourcesProcessesControlling2015]. Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Given the structure of equation 1, calculations of this flux are highly sensitive to our measurements/estimates of *k*. Note that *k* is different for different gases and water temperatures [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

$$Flux=k(C_{water}-C_{air}) \mathbf{(1)}$$

*k* is also vital to understanding freshwater ecology through bulk river metabolism [gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018]. The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *D* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature. When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the fundamental biogeochemical processes that occur in river networks. Therefore, both closing the global carbon cycle and further understanding freshwater ecology rely on accurate and high-fidelity assessments of *k*.
$$\frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water}) \mathbf{(2)}$$
Researchers use a range of methods to measure *k*, from floating domes to eddy covariance towers to measure at a specific point in the river. They also inject tracer gases to estimate reach-scale gas exchange [@hallGasExchangeStreams2020]. These approaches are ideal but are infeasible when working at the river network scale across potentially tens of thousands of rivers. So, predictive models are used to estimate *k* from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. These models often, though not alwyas, predict $k_{600}$, or *k* that has been normalized to a Schmidt number of 600 ($CO_2$ at 20 degrees Celsius or $O_2$ at 17.5 degrees).

There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@hallGasExchangeStreams2020] and extending the original work of @streeterMeasuresNaturalOxidation1935 and @phelpsStudiesSelfPurificationStreams1914. The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities. Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between *k* and hydraulic properties. However, many of them are not based on fundamental mass transport theory and rather are simple empirical relations. This means that those models' parameters are sensitive to the training data used [@wangPhysicallyBasedScaling2021].

Multiple predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018]. This is convenient because both *eD* and $U_\star$ are calculated using easily estimated hydraulic parameters: $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976] and $U_\star=\sqrt{gS_hR_h}$, where *g* is gravitational acceleration $[L/T^2]$, $S_h$ is water surface slope $[L/L]$, *V* is average flow velocity $[L/T]$ and $R_h$ is the hydraulic radius [L] (which often approximates *D*). Because slope is readily available in any hydrographic data product, most efforts to predict *k* across thousands of rivers are thereby limited by the quality of the final parameters *V* or $R_h$. Since both must be measured in the field or themselves estimated by other empirical methods, uncertainty in *V* or $R_h$ must dominate any errors in *k* estimates. This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins and across thousands of rivers, researchers have used 'global-scope' hydraulic geometry (HG) models that extend HG beyond its traditional use. Traditional HG models are simple power law relations between streamflow and channel width, depth, or velocity at a specific cross-section in a river [@leopoldHydraulicGeometryStream1953a] and a large body of geomorphology work has attempted to parse out process-based explanations for observed HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a]. This has been extended to 'downstream HG' which relates a characteristic streamflow (i.e. mean annual or bankful discharge) to characterisitc hydraulics along a specific river's course [@leopoldHydraulicGeometryStream1953a].

For the purposes of predicting *k* from *V* or $R_h$, biogeochemists have further extended HG models by training them on large hydraulics datasets for thousands of rivers and under all streamflow conditions [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (traditional HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977]. This means that the 'global-scope' HG models that are usually used are highly reliant on their training data and can potentially yield pretty different hydraulics for the same streamflow. It is currently not well understood how sensitive global estimates of riverine gas evasion are to the specific 'global-scope' HG model that is employed by the worker.

@@

SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. This yields an average sampling resolution of 11 days. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged methods are termed McFLIs or 'Mass Conserved Flow Law Inversion' algorithms [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a]. These use basic geomorphic theories and the concept of 'prior river knowledge' to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. McFLIs are readily implemented in any river that SWOT can observe and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020a; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation of McFLIs into a traditional hydrologic model [@ishitsukaCombiningOpticalRemote2020].

While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate *k*, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems. Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and RS of *k*. Bayesian techniques have been previously used to concurrently solve for *k* and *M* in equation 2 [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While these studies are field based, require high-fidelity in situ data, and are not applicable to *k* estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing *k* when parameter equifinality is a problem. This matches McFLI logic [@andreadisConstrainingAssimilationSWOT2020d].

In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate *k*. This would in turn improve our understanding of riverine gas fluxes in ungauged basins (equation 1), as well as potentially help in parameterizing stream metabolism models (equation 2). Therefore, this manuscript aims to answer two questions:

@@

To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{o_2}$ estimates and its explicit Bayesian uncertainity that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{o_2}$ algorithm BIKER, or the 'Bayesian Inversion of the $k_{o_2}$ Evasion Rate'. We validate BIKER on simulated SWOT data for 49 SWOT-observable rivers from around the world, using hydraulic models  to produce SWOT-like data as SWOT has not yet launched. This is standard practice in the SWOT community [@frassonExploringFactorsControlling2021; @durandIntercomparisonRemoteSensing2016b]. We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, SWOT is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 49 rivers as calculated using BIKER and previously published 'global-scope' HG models to compare BIKER with established in situ techniques.

##2 Methods
To build BIKER, we join a process-based model for $k_{o_2}$ (section 2.1) with a McFLI framework for inverting SWOT measurements (section 2.2). Following the description of that process, we describe the validation (section 2.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 2.4). A flowchart detailing the entire study is provided as Figure 1.

![Figure 1. Flowchart of methodology used in this study. We begin by choosing a process-based model for k (section 2.1) that is a function of river hydraulics measurable by SWOT. Then, we implement this model within a McFLI ('BIKER') to estimate k solely from river width and water surface slope (section 2.2). Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 2.4). See section 2.3 for the validation setup.](cache/flowchart.jpg)

###2.1 Choosing a predictive model for *k*
To predict k from just the SWOT observables,river width *W* and water surface elevation *H*, a physical model for *k* is first selected. As mentioned in section 1, there are nearly two dozen predictive models for *k* that have been devolped since the 1950s [@wangPhysicallyBasedScaling2021; @hallGasExchangeStreams2020]. Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021. They significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using inverse modeling of DO timeseries. This approach was also briefly mentioned in section 1. These methods calibrate $k_{o_2}$, along with *M* (discretized as seperate respiration and gross primary production terms), to high-fidelity, in situ datasets of DO with the goal of minimizing the difference between the observed DO and the modeled DO [e.g. @applingOvercomingEquifinalityLeveraging2018; @chapraDeltaMethodEstimating1991]. @wangPhysicallyBasedScaling2021 used the *streamMetabolizer* technique [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States and showed that $k_{o_2}$ models deduced from the classic, process-based 'thin-film' and 'surface renewal' theories for mass transport yield approximately identical parameters when fit to field measurements or this new dataset of simulated $k_{o_2}$. They also showed through cross-validation that these models are robust to overfitting on specific sets of data (a concern with many of the aforementioned predictive models- section 1). This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used. See @hallGasExchangeStreams2020 and @wangPhysicallyBasedScaling2021 for more in-depth explanations of the thin-film and surface renewal theories with respect to rivers. In short, these are the standard physical models for conceptualizing gas exchange in various aquatic environments from lakes to estuaries to rivers. The better fit, and more parsimonious, of the two models from @wangPhysicallyBasedScaling2021 is reprinted as equation 3(with reported coefficients of determination $r^2$) where $\alpha$ is the fitted parameter in the linear regression. Equation 3, 2nd line was fit to 588 field measurements of $k_{o_2}$ while equation 3, third line was fit to 3,919 simulated $k_{o_2}$ values at 35 rivers.
$$\begin{eqnarray} k_{o_2} &=& \alpha U_\star \mathbf{(3)} \\ &=& 47.67U_\star (r^2:0.53) \\ &=& 49.71U_\star (r^2: 0.76)  \end{eqnarray}$$

We implement this model within BIKER (using 48 as a reasonable value for $\alpha$ given equation 3) for a few reasons. First, as suggested earlier @wangPhysicallyBasedScaling2021 provides compelling evidence that an $\alpha$ of approximately 48 is uniform across rivers and streams, regardless of the training data used. This is encouraging for use within BIKER as BIKER is specifically designed to be as river-agnostic as possible and so we sought a fully generalized equation. Second, it yields a simple linear relationship across all rivers, regardless of their size or steepeness. This is not necessairly true of other *k* models, for example those based on *eD* (section 1). Recent work has shown that *k* does not scale with *eD* via the same statistical parameters across all river geomorphologies [@ulsethDistinctAirWater2019a]. We further explored the relationship presented in that paper and found that the $k \sim eD$ relationship fundamentally breaks down in rivers with low *eD*, which corresponds to nearly all rivers that SWOT will observe (**supp figure, need to do**). Therefore, an *eD* based predictive model is less useful in SWOT rivers. Third, we sought to explicitly predict $k_{o_2}$, as its use extends beyond constraining gas fluxes from rivers and into parameterizing river metabolism models (equation 2).

We stress that most other predictive models for *k* could be implemented within BIKER. However, our goal of using BIKER on any river that is observable by SWOT necessitates that we pick an equation that is applicable across as many rivers as possible. Future work should look at using a river-specific *k* equation to pseudo-calibrate BIKER to a specific river.

###2.2 Developing BIKER
With a predictive model for $k_{o_2}$ chosen, we now implement it within the McFLI framework. The approach used here is informed by the @hagemannBAMBayesianAMHGManning2017a McFLI algorithm for ungauged RSQ, further explored in more recent work [@brinkerhoffConstrainingRemoteRiver2020a]. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{o_2}$ as a Bayesian remote sensing problem that can be solved using SWOT data.

BIKER and Bayesian inference in general starts from Bayes rule, after acknowleding that it needs only to be specified as a proportionality to approximate the full posterior distribution (equation 4). In equation 4, $\Theta$ is some set of parameters we want to solve for (including $k_{o_2}$), *x* is the observed data, $f(x|\Theta)$ is the 'likelihood function' or sampling model conditional on the parameters, and $p(\Theta)$ is the joint prior distribution of the parameters. For BIKER, *x* is the SWOT-observables *W* and *H*. First, we detail the likelihood specification and then the prior specifications.

$$p(\Theta|x) \propto f(x|\Theta)p(\Theta) \mathbf{(4)}$$

####2.2.1 BIKER likelihood
To concieve of equation 3 as a likelihood function, $k_{o_2}$ first needs to be written as a function of SWOT-observables *W* and *H* (which provides water surface slope $S_h$). This algebra is carried out in equation 5, where we assume that $R_h \approx \frac{A}{W}$ where *A* is wetted channel area. *A* is further split into the SWOT-observable portion *dA* and the unobservable portion $A_0$ following @durandEstimatingReachaveragedDischarge2014a. *dA* is estimated by assuming a rectangular river channel so that $dA \approx \delta H*W$.

>>>>

---
title: "Remote sensing riverine gas exchange and carbon efflux in ungauged rivers from SWOT observations"
author: 'CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##1 Introduction
**I don't like this opening paragraph** Vital to sustaining life, the Earth's carbon cycle moves carbon between the atmosphere, geosphere, hydrosphere, and biosphere via a complex set of interconnected biogeochemical processes. The primary connection between the terrestrial and oceanic biospheres is the global river network, which exports much carbon from the continents to the oceans [@colePlumbingGlobalCarbon2007a]. However en route to the oceans, this river network additionally transports and transforms sediments, nutrients, organic matter, dissolved gases, and other constituents. Organic matter and inorganic nutrients are also produced and/or consumed within rivers via ecosystem metabolism [@odumPrimaryProductionFlowing1956]. One net effect of all of this is the global river network is usually supersaturated with dissolved gases and evades substantial amounts of carbon from its water surface to the atmosphere [@colePlumbingGlobalCarbon2007a].

Current estimates of total carbon dioxide ($CO_2$) evasion to the atmosphere vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a], with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Further, in-river metabolism is increasingly understood to contribute to this flux [@hotchkissSourcesProcessesControlling2015]. Equation 1 represents this flux for any sparingly soluble gas $[M/L^2T]$ given the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Given the structure of equation 1, calculations of this flux are very sensitive to one's measurements/estimates of *k*. Note that *k* is different for different gases and water temperatures [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

$$\mathbf{(1)} Flux=k(C_{water}-C_{air})$$

*k* is also vital to understanding freshwater ecology through bulk river metabolism [more specifically, gross primary production and ecological respiration- @bernhardtMetabolicRegimesFlowing2018]. The general mass-balance equation for river metabolism (*M* $[M/L^2T]$) is provided as equation 2, where *DO* is the dissolved oxygen concentration $[M/L^3]$, *D* is river depth [L], and $k_{o_2}$ is *k* for oxygen at some temperature. When viewing equations 1 and 2 in aggregate, it is clear that *k* plays a critical role in two of the core, fundamental biogeochemical processes that occur in river networks. Both are reliant on accurate and high-fidelity assessments of *k*, as well as a rigorous understanding of the physical mechanisms driving *k*.
$$\mathbf{(2)} \frac{dDO}{dt}=\frac{M}{D}+\frac{k_{o_2}}{D}(DO_{air}-DO_{water})$$
Researchers use a range of methods to measure *k*, from floating domes to eddy covariance towers to measure at a specific point in the river. They also inject tracer gases to estimate reach-scale gas exchange [@hallGasExchangeStreams2020]. These approaches are ideal but are infeasible when working at the river network scale across potentially tens of thousands of rivers. So, predictive models are used to estimate *k* from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. These models often, though not always, predict $k_{600}$ or *k* that has been normalized to a Schmidt number of 600 ($CO_2$ at 20 degrees Celsius or $O_2$ at 17.5 degrees).

There is a long tradition of water quality engineers and biogeochemists developing predictive models for different forms of *k*, going back to at least the 1950s [@hallGasExchangeStreams2020] and extending the original work of @streeterMeasuresNaturalOxidation1935 and @phelpsStudiesSelfPurificationStreams1914. The goals of these models vary, from predicting oxygen reaeration coefficients to predicting $CO_2$ gas transfer velocities. Most of these models operate under the basic assumption that river channel hydraulics govern gas exchange and so simple statisical relationships can be generated between *k* and hydraulic properties. However, many of them are not based on fundamental mass transport theory and rather are simple empirical relations. This means that those models' parameter values are sensitive to the training data used [@wangPhysicallyBasedScaling2021].

Many of these predictive models are based on the observation that *k* is generally correlated with both the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a] and the water surface surface shear velocity $U_\star [L/T]$ [@wangPhysicallyBasedScaling2021; @katulStructureFunctionModel2018]. This is convenient because both *eD* and $U_\star$ are calculated using easily estimated hydraulic parameters: $eD \approx gS_hV$ [@tsivoglouTracerMeasurementReaeration1976] and $U_\star=\sqrt{gS_hR_h}$, where *g* is gravitational acceleration $[L/T^2]$, $S_h$ is water surface slope $[L/L]$, *V* is average flow velocity $[L/T]$ and $R_h$ is the hydraulic radius [L] (which often approximates *D*). Because slope is readily available in any hydrographic data product, most efforts to predict *k* across thousands of rivers are thereby limited by the quality of the final parameters *V* or $R_h$. Since both must be measured in the field or themselves estimated by empirical models, uncertainty in *V* or $R_h$ dominates errors in *k* estimates. This is exacerbated in ungauged basins that cover large areas, especially in the carbon-rich Arctic inland waters, as no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins and across thousands of rivers, researchers have used 'global-scope' hydraulic geometry (HG) models that extend HG beyond its 'at-a-station' use. At-a-station HG models are simple power law relations between streamflow and channel width, depth, or velocity at a specific cross-section in a river [@leopoldHydraulicGeometryStream1953a] and a large body of geomorphology work has attempted to parse out process-based explanations for at-a-station HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a]. This has been extended to 'downstream HG' which relates a characteristic streamflow (i.e. mean annual or bankful discharge) to characteristic hydraulics along a specific river's course [@leopoldHydraulicGeometryStream1953a].

For the purposes of predicting *k* from *V* or $R_h$, biogeochemists have further extended HG models by training them on large hydraulics datasets for thousands of rivers and under all streamflow conditions [e.g @borgesGloballySignificantGreenhousegas2015a; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters are either cross-section specific and highly variable from river to river (at-a-station HG) or they are discharge specific [downstream HG- @gleasonHydraulicGeometryNatural2015; @parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977]. This suggests that the 'global-scope' HG models that are usually used are also sensitive to their training data and can potentially yield pretty different hydraulics for the same streamflow. It is currently not well understood how sensitive global estimates of riverine gas evasion are to the specific 'global-scope' HG model that is employed by the worker.

@@

SWOT is expected to launch in 2022 and provide the world's first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. This yields an average sampling resolution of 11 days. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. A decade of SWOT work has explored the multi-parameter problem of estimating the river hydraulic parameters of roughness and bathymetry from remote sensing to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @brissetAssimilationAltimetricData2018; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged methods are termed McFLIs or 'Mass Conserved Flow Law Inversion' algorithms [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020a; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a]. These use basic geomorphic theories and the concept of 'prior river knowledge' to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. McFLIs are readily implemented in any river that SWOT can observe and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020a; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information via data assimilation of McFLIs into a traditional hydrologic model [@ishitsukaCombiningOpticalRemote2020].

While all McFLIs to date have been developed in the context of RSQ, there is no reason their logic cannot be used to estimate *k*, particularly as McFLIs often employ Bayesian inference for equifinal inverse problems. Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015], and this problem is common to both RSQ and RS of *k*. Bayesian techniques have been previously used to concurrently solve for *k* and *M* for different forms of equation 2 [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While these studies require high-fidelity in situ data and are not applicable to *k* estimation in ungauged rivers, they suggest that Bayesian techniques could be useful for remotely sensing *k* when parameter equifinality is a problem. This matches McFLI logic [@andreadisConstrainingAssimilationSWOT2020d].

In this context, we hypothesize that a combination of SWOT data, a Bayesian McFLI paradigm, and gas exchange theory can successfully estimate *k* using absolutely no in situ information. This would in turn improve our understanding of riverine gas fluxes in ungauged basins (equation 1), as well as potentially help in parameterizing stream metabolism models (equation 2). Therefore, this manuscript aims to answer two questions:

@@

To answer these questions, we developed a new McFLI algorithm that ingests SWOT data and produces $k_{o_2}$ estimates and their explicit Bayesian uncertainities that requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{o_2}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{o_2}$ Evasion Rate'. We validate BIKER on simulated SWOT data for 49 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data as SWOT has not yet launched. This is standard practice in the SWOT community [@frassonExploringFactorsControlling2021; @durandIntercomparisonRemoteSensing2016b]. We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016], as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypotetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 49 rivers as calculated using BIKER and previously published in situ techniques.

##2 Data
Numerous datasets were used in this study to develop and validate BIKER and are detailed below.

BIKER validation (section 3.3) was performed on 49 SWOT-simulated rivers. Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on SWOT-like data. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as RS observations and are used as the sole inputs to BIKER. Here, we use 49/51 rivers collected by @frassonExploringFactorsControlling2021 and @durandIntercomparisonRemoteSensing2016b. These are the two benchmarking studies that have explored RSQ algorithm performance for the SWOT mission. Please consult both of those papers for all of the rivers' locations as well as hydraulic model specifications. Ultimately, the 49 rivers are spread across the United States, France, Italy, the United Kingdom, and Bangladesh. We omit both models for the Saint Lawrence River, Canada from @durandIntercomparisonRemoteSensing2016b because they lack enough hydraulic information to calculate $U_\star$.

We also assess the influence of measurement error on BIKER's performance. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error and width measurement error. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, leading to taller landscape features appearing closer to the sensor than shorter landscape features that are the same horizontal distance from the sensor. For river heights and slopes, we use the error model devolped by @durandHowWillRadar2020a and implemented on 17/49 of the rivers by @frassonExploringFactorsControlling2021. Width errors were derived using a model built by @frassonExploringFactorsControlling2021 based on SWOT-simulator runs for the Sacramento river, California [@frassonAutomatedRiverReach2017b] and the Po river, Italy [@domeneghettiCharacterizingWaterSurface2018]. Those two simulations allowed for explicit characterization of width errors that were then extrapolated onto the other 15 rivers. Width errors due to poor water classification are ignored as they were in @frassonExploringFactorsControlling2021. Please consult @frassonExploringFactorsControlling2021 for the specifics of how these realistic errors were incorporated into the hydraulic models.

BIKER prior specifications (section 3.2.2) require the use of a training dataset of field-measured river hydraulics. To do this, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of average channel velocity, width, depth, and discharge from across the continental United States.

For the $CO_2$ evasion and carbon efflux calculations (section 3.4), we use 26 bi-weekly dissolved $CO_2$ samples made by @beaulieuControlsGasTransfer2012a at one location in the Ohio River for one calender year from 2008-2009 (Figure S1). Note that this $CO_2$ data is for the Ohio River only but was applied to all 49 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. **Kelly just published hourly CO2 data for the CT, maybe we should use that instead to get mean daily values?**

Finally, we also made extensive use of data and models developed by workers to predict gas evasion rates from river hydraulics. Specifically, we used the data measured, collected, or simulated by @raymondScalingGasTransfer2012a, @ulsethDistinctAirWater2019a, and @wangPhysicallyBasedScaling2021.

##3 Methods
To build BIKER, we join a process-based model for $k_{o_2}$ (section 3.1) with a McFLI framework for inverting SWOT measurements via Bayesian inference (section 3.2). Following the description of that process, we describe the validation setup (section 3.3) and the workflow for comparing estimated bulk carbon effluxes from a suite of 'global-scope' HG models (section 3.4). A flowchart detailing the entire study is provided as Figure 1.

![Figure 1. Flowchart of methodology used in this study. We begin by choosing a process-based model for k (section 3.1) that is a function of river hydraulics measurable by SWOT. Then, we implement this model within a McFLI ('BIKER') to estimate k solely from river width and water surface slope (section 3.2). Finally, we couple BIKER with in situ CO2 data to compare BIKER-estimated gas fluxes with established in situ methods (section 3.4). See section 3.3 for the validation setup.](cache/flowchart.jpg)

###3.1 Choosing a predictive model for *k*
To predict *k* from just the SWOT observables, a physical model for *k* is first selected. As mentioned in section 1, there are nearly two dozen predictive models for *k* that have been devolped since the 1950s [@wangPhysicallyBasedScaling2021; @hallGasExchangeStreams2020]. Recently, many of these models were reexplored in rivers and streams by @wangPhysicallyBasedScaling2021, who significantly expanded the existing training datasets of field-measured $k_{o_2}$ by using the *streamMetabolizer* model to calibrate a form of equation 2 to high-fidelity in situ DO datasets [@applingOvercomingEquifinalityLeveraging2018] at 35 rivers across the United States. They showed that two $k_{o_2}$ models deduced from classic, process-based theories for mass transport yield approximately identical parameters when fit to ethier field measurements or this new dataset of simulated $k_{o_2}$. They also showed through cross-validation that these models are robust to overfitting on specific sets of data (a concern with many of the aforementioned predictive models- section 1). This suggests that there are uniform scaling realtionships between certain hydraulic properties and $k_{o_2}$ that exist regardless of the training data used. @wangPhysicallyBasedScaling2021's better fit, and more parsimonious, of the two models is reprinted as equation 3 (with that paper's reported coefficients of determination $r^2$). $\alpha$ is the fitted parameter in the linear regression. Equation 3-2nd line was fit to 588 field measurements of $k_{o_2}$ while equation 3-third line was fit to 3,919 simulated $k_{o_2}$ values at 35 rivers.
$$\mathbf{(3)} \begin{eqnarray} k_{o_2} &=& \alpha U_\star \\ &=& 47.67U_\star (r^2:0.53, measured) \\ &=& 49.71U_\star (r^2: 0.76, simulated)  \end{eqnarray}$$
